Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2018-09-14 Manuscript/manuscript.tex
Comments added into the manuscript

\todo{Joe: It might be a very short section -- Shall we merge it with the simulation details section?}

@@

The structures of lipid bilayers in simulations without ions were validated against NMR 
by calculating the order parameters for the C-H bonds and against \mbox{X-ray} scattering experiments by evaluating the scattering form factors
as was done in Refs.~\citenum{catte16, melcr18, nmrlipids_proj4}.
%NMR order parameters validate the structures sampled by the individual lipid molecules with atomic resolution. 
The simulated order parameters were calculated for all C-H bonds in lipid molecules, 
while all order paremeters in the head group and only some in the \ce{C-H} groups in tails were resolved experimentally
(see next section and Fig.~\ref{simVSexpNOions_POPS}).

@@

We used the SPC/E~\cite{Berendsen1987} water model 
and ECC-ions \cite{martinek17, kohagen16, Pluharova2014} 
for the simulations with ECC-lipids.
Force field parameters for POPC were taken from Ref.~\citenum{melcr18}. 
The ECC-POPS model as derived again from the Lipid14 force field~\cite{dickson14, lipid17-future}, as was described above.
For a direct comparison, 
we also performed simulations with the Lipid14 model~\cite{dickson14},
the TIP3P water model~\cite{jorgensen83}, 
and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006}.

@@

\begin{table*}[tbp]
\centering
\caption{  List of MD simulations with
           numbers of individual molecules for each simulation
           at temperature $T$. 
        }
\label{tbl:sim-list}
\begin{tabular}{l c c r r r r r }
  lipids/counter-ions &  \ce{K^+}  &  \ce{Na^+} & \ce{Ca^{2+}} & \ce{Cl^-} & \ce{H2O} &  PC:PS total  &  $T$~(K)  \\
  \hline
PC:PS (1:1)/Na+  &  0  &  64  &  0  &  0  &  5253  &  64:64  &  298 \\ 
PC:PS (1:1)/K+  &  64  &  0  &  0  &  0  &  5253  &  64:64  &  298 \\ 
PC:PS (10:1)/Na+  &  0  &  6  &  0  &  0  &  3600  &  60:6  &  298 \\ 
PC:PS (10:1)/K+  &  6  &  0  &  0  &  0  &  3600  &  60:6  &  298 \\ 
PC:PS (4:1)/Na+  &  0  &  12  &  0  &  0  &  3600  &  48:12  &  298 \\ 
PC:PS (4:1)/K+  &  12  &  0  &  0  &  0  &  3600  &  48:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  13  &  26  &  3561  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  26  &  52  &  3522  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  0  &  0  &  3600  &  60:12  &  298 \\ 
PC:PS (5:1)/K+  &  12  &  0  &  0  &  0  &  3600  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  39  &  78  &  3483  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  39  &  12  &  0  &  39  &  3483  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  51  &  0  &  39  &  3483  &  60:12  &  298 \\ 
  \hline
PS/K+  &  72  &  0  &  0  &  0  &  3600  &  0:72  &  298 \\ 
PS/Na+  &  0  &  72  &  0  &  0  &  3600  &  0:72  &  298 \\ 
  \hline
\end{tabular}
\end{table*}

@@

@@

\begin{figure}[htb!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    Changes of the head group order parameters of a POPC:POPS (5:1) bilayer as a function of PS content
    with \ce{Na^+} and \ce{K^+} counterions from simulations with ECC-lipids and Lipid17 \cite{lipid17-future} 
    and experiments (only \ce{Na^+} counterions). \cite{roux90}. 
    The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    where the response is closer to the experiment (which uses only \ce{Na^+} counterions). 
    Experimental points for POPS inherently contain larger systematic error, 
    as they come from independent experiments.
  } 
\end{figure}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{NaCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K.
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.}
\end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{KCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields and experiments at 298 K. \citep{roux90}
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.}
\end{figure}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_na_k_cl_wat_phos_models-compar_4-6_NaCl-and-KCl-series.pdf}
  \caption{\label{fig:nacl-dens_PCPS} 
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} and \ce{Cl^-} along the membrane normal axis 
    for the negatively charged membranes with the composition of 5\,PC:1\,PS using ECC-lipids and ECC-ions.  
    The top profile shows the simulation without any additional salt concentration, i.e., only with \ce{Na^+} counterions. 
    The middle profile shows the simulation with an additional \ce{KCl} concentration and \ce{Na^+} counterions. 
    The bottom profile shows the simulation with an additional \ce{NaCl} concentration and \ce{Na^+} counterions, which are not distinguished from the added salt. 
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    the density profiles of~\ce{Cl^-}, \ce{K^+} and \ce{Na^+} ions are divided by 2, and 
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively.  
    } 
\end{figure}

The difference between the affinities of \ce{Na^+} and \ce{K^+} to neutral and negatively charged membranes
can be described by their relative surface excesses with respect to water, $\Gamma ^{w} _{ion}$, 
which are shown in the plots of the density profiles of the ions in Figs.~\ref{fig:nacl-dens} (neutral bilayer) and \ref{fig:nacl-dens_PCPS} (negatively charged bilayer). 
Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
While \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged bilayer,
the value of $\Gamma^{w}_{Na}$ for \ce{Na^+} changes from negative to positive
in a neutral POPC bilayer versus in a negatively charged bilayer with the composition 5\,PC:1\,PS.
%This means that at the given concentration the bilayer interface has a small preference to \ce{Na^+} cations compared to water molecules. 
Interestingly, this value is slightly decreased in the presence of an additional \ce{NaCl} concentration adding also \ce{Cl^-} anions, 
which are not present in the system when only counterions are used
(bottom resp. top plot in Fig.~\ref{fig:nacl-dens_PCPS}). 
The relative surface excess of calcium, $\Gamma^{w}_{Ca}$, also reflects
the higher affinity of the negatively charged PC:PS (5:1) membrane compared to a neutral POPC bilayer; 
its value for the latter is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$, 
while for the PC:PS mixture it is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable concentrations $C _{Ca} = 260\pm 30 \mathrm{mM} $.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_cacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_cacl.pdf} 
  \caption{\label{fig:delta_ordPar_CaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{CaCl2} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields and experiments at 298~K. \citep{roux90}
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.}
\end{figure}

@@

\begin{figure}[htbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_ca_na_cl_wat_phos_models-compar_1-3_CaCl2-series.pdf}
  \caption{\label{fig:cacl-dens_PCPS} 
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along the normal of the membrane starting at the centre
    for the negatively charged membrane with the composition 5\,PC:1\,PS
    at various bulk concentrations of \ce{CaCl2} from simulations with ECC-lipids. 
    All profiles contain \ce{Na^+} counterions and an additional concentration of \ce{CaCl2}. 
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}},  
    the density profiles of~\ce{Cl^-} ions are divided by 2, and 
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively.  
    } 
\end{figure}

@@

%%% discuss the table with Gammas and follow up with molecular details, where do cations bind? Support with the spatial density figure
The density profiles of ions suggest that
the dominant contribution to the binding of \ce{Ca^{2+}} to phospholipid bilayers
comes from the interactions with the phosphate groups in both POPC and POPS. 
%The population analysis also suggests 
%that the calcium cations prefer to reside in the phosphate region of the membrane. 
%Such a finding is further validated with a more detailed analysis using Markov state modeling (MSM) \citep{Pande_MSM_paper} \todoi{Add papers reviewing MSM, e.g. recent Pande's paper.}. 
%The set of states of a calcium cation 
%encompassed all possible combinations of up to three surrounding lipids. 
%In the case of PC, we used only the phosphate moitety, which forms the dominant contribution for calcium binding.
%For PS we also distinguished configurations in which calcium interacts with the carboxylate moiety. 
%All possible combinations of such states were used to build a Markov model at a lag time $25\,\mathrm{ns}$ using pyEMMA code by \citet{pyemma} \todoi{add citation for pyemma}. 
%The resulting MSM was validated using Chapman-Kolmogorov test \citep{FrankNoe_papers_MSM} \todoi{Add citation for Noe's papers on MSM, especially Chapman Kolmogorov test.}. 
%The stationary distribution of the states reveals a strong preference of the calcium cations to reside in the phosphate region of the mixed PC-PS membrane. 
%When interacting with PS, configurations with the phosphate moiety from either PC or PS dominate the population.
%Moreover, states containing interactions with the carboxylate group in PS 
%bear higher probability when interacting also with the phosphate group of the same lipid or other lipids. 
This is also reflected in the shift of the mean orientation 
of the \ce{COO^-} group from $62^\circ$ to $73^\circ$ (420~mM \ce{CaCl2})
which was measured as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid. 
The interactions of the carboxylate group in PS with calcium and other phosphate groups
shed light into the qualitatively different response of the head group order parameters $\alpha$ and $\beta$ in PS compared to PC 
(see Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_CaCl_PCPS}). 
We find that the complex response of the head group order parameters of POPS 
is affected by the conformational changes of the carboxylate group,
which is attracted more towards the phosphate region, 
where the calcium cations dominantly bind. 
In line with the experimental work by \citet{browning80},
this is also very likely the reason, 
why the magnitude of the P-N vector change in POPS is diminished compared to POPC, 
which is not restrained by an additional cation binding group like \ce{COO^-} in POPS.

>>>>

\todo{Joe: It might be a very short section -- Shall we merge it with the simulation details section?
SAMULI: Let's write it first and then see.}

@@

\todo{SAMULI: Short discussion about mixtures should be added by citing the NMRlipids IV.}
The structures of lipid bilayers in simulations without ions were validated against NMR 
by calculating the order parameters for the C-H bonds and against \mbox{X-ray} scattering experiments by evaluating the scattering form factors
as was done in Refs.~\citenum{catte16, melcr18, nmrlipids_proj4}.
%NMR order parameters validate the structures sampled by the individual lipid molecules with atomic resolution. 
The simulated order parameters were calculated for all C-H bonds in lipid molecules, 
while all order paremeters in the head group and only some in the \ce{C-H} groups in tails were resolved experimentally
(see next section and Fig.~\ref{simVSexpNOions_POPS}).

@@

We used the SPC/E~\cite{Berendsen1987} water model 
and ECC-ions \cite{martinek17, kohagen16, Pluharova2014} 
for the simulations with ECC-lipids.
Force field parameters for POPC were taken from Ref.~\citenum{melcr18}. 
The ECC-POPS model as derived again from the Lipid14 force field~\cite{dickson14, lipid17-future}, as was described above.
For a direct comparison, 
we also performed simulations with the Lipid14 model~\cite{dickson14},
the TIP3P water model~\cite{jorgensen83}, 
and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006}.
\todo{The usage of Dang ions should be justified here.}

@@

\begin{table*}[tbp]
\centering
\caption{  List of MD simulations with
           numbers of individual molecules for each simulation
           at temperature $T$. 
         }
         \todo{SAMULI: Temperature column is not necessary because it is always the same.}
         \todo{SAMULI: Columns reporting the additional ion concentrations would be useful.}
         \todo{SAMULI: Order of the PC:PS (5:1)/Na+ systems in little bit confusing.}
\label{tbl:sim-list}
\begin{tabular}{l c c r r r r r }
  lipids/counter-ions &  \ce{K^+}  &  \ce{Na^+} & \ce{Ca^{2+}} & \ce{Cl^-} & \ce{H2O} &  PC:PS total  &  $T$~(K)  \\
  \hline
PC:PS (1:1)/Na+  &  0  &  64  &  0  &  0  &  5253  &  64:64  &  298 \\ 
PC:PS (1:1)/K+  &  64  &  0  &  0  &  0  &  5253  &  64:64  &  298 \\ 
PC:PS (10:1)/Na+  &  0  &  6  &  0  &  0  &  3600  &  60:6  &  298 \\ 
PC:PS (10:1)/K+  &  6  &  0  &  0  &  0  &  3600  &  60:6  &  298 \\ 
PC:PS (4:1)/Na+  &  0  &  12  &  0  &  0  &  3600  &  48:12  &  298 \\ 
PC:PS (4:1)/K+  &  12  &  0  &  0  &  0  &  3600  &  48:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  13  &  26  &  3561  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  26  &  52  &  3522  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  0  &  0  &  3600  &  60:12  &  298 \\ 
PC:PS (5:1)/K+  &  12  &  0  &  0  &  0  &  3600  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  12  &  39  &  78  &  3483  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  39  &  12  &  0  &  39  &  3483  &  60:12  &  298 \\ 
PC:PS (5:1)/Na+  &  0  &  51  &  0  &  39  &  3483  &  60:12  &  298 \\ 
  \hline
PS/K+  &  72  &  0  &  0  &  0  &  3600  &  0:72  &  298 \\ 
PS/Na+  &  0  &  72  &  0  &  0  &  3600  &  0:72  &  298 \\ 
  \hline
\end{tabular}
\end{table*}

@@

\todo{SAMULI: Lipid17 simulations are too compressed, which is probably due to the
  overbinding of Dang ions. Lipid17 with Åqvist ions (default in Amber) has larger
  area in NMRlipids IV. This is similar to ECC-POPC paper where changing from Åqvist to
  Dang or ECC-ions increased binding affinity in Lipid14. However, the headgroup order
  parameters are better in ECC-POPS compared to Lipid17 simulations with any ions.
  These should be discussed in this section.
}

@@

\begin{figure*}[!htb] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[width=\figwidthsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    Changes of the head group order parameters of a POPC:POPS (5:1) bilayer as a function of PS content
    with \ce{Na^+} and \ce{K^+} counterions from simulations with ECC-lipids and Lipid17 \cite{lipid17-future} 
    and experiments (only \ce{Na^+} counterions). \cite{roux90}. 
    The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    where the response is closer to the experiment (which uses only \ce{Na^+} counterions). 
    Experimental points for POPS inherently contain larger systematic error, 
    as they come from independent experiments.
  }
  \todo{SAMULI: Caption says Åqvist, but I guess that Dang is used also here?} \\
  \todo{SAMULI: Do you have a simulation of Lipid14 pure POPC at 298K?
    This is not in the table. This simulation would be useful also for NMRlipids IV.
  } \\
  \todo{SAMULI: Error bars for lipid14 missing.}
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{NaCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields at 298 K.
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.} \\
  \todo{SAMULI: Ion concentrations should be updated to the concentrations in water before solvating the lipids.} \\
  \todo{SAMULI: Error bars should be recalculated with the fixed code} \\
  \todo{SAMULI: We should have at least one more points with higher concentration (maybe close to 2M), or maybe two more points.} \\
  \end{figure}

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_kcl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_KCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{KCl} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields and experiments at 298 K. \citep{roux90}
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.} \\              
  \todo{SAMULI: Ion concentrations should be updated to the concentrations in water before solvating the lipids.} \\
  \todo{SAMULI: Error bars should be recalculated with the fixed code} \\
  \todo{SAMULI: We should have at least one more points with higher concentration (maybe close to 2M), or maybe two more points.} \\
\end{figure}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_na_k_cl_wat_phos_models-compar_4-6_NaCl-and-KCl-series.pdf}
  \caption{\label{fig:nacl-dens_PCPS} 
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} and \ce{Cl^-} along the membrane normal axis 
    for the negatively charged membranes with the composition of 5\,PC:1\,PS using ECC-lipids and ECC-ions.  
    The top profile shows the simulation without any additional salt concentration, i.e., only with \ce{Na^+} counterions. 
    The middle profile shows the simulation with an additional \ce{KCl} concentration and \ce{Na^+} counterions. 
    The bottom profile shows the simulation with an additional \ce{NaCl} concentration and \ce{Na^+} counterions, which are not distinguished from the added salt. 
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    the density profiles of~\ce{Cl^-}, \ce{K^+} and \ce{Na^+} ions are divided by 2, and 
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively.  
  }
  \todo{SAMULI: Top figure should go together with figure of only K+ counterions and figure with pure PS mentioned
    another todo point
  } \\
  \todo{SAMULI: Division of ion densities with 2 in figure without calcium is not necessary and makes
  comparison to other results more difficult.}
\end{figure}

The difference between the affinities of \ce{Na^+} and \ce{K^+} to neutral and negatively charged membranes
can be described by their relative surface excesses with respect to water, $\Gamma ^{w} _{ion}$, 
which are shown in the plots of the density profiles of the ions in Figs.~\ref{fig:nacl-dens} (neutral bilayer)
\todo{SAMULI: We should have a figure showing density profiles of Na+ and K+ in pure POPS and PC:PS (5:1) mixture
  with only counterions from ECC-POPS model and Lipid17 model.
}
and \ref{fig:nacl-dens_PCPS} (negatively charged bilayer). 
Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
While \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged bilayer,
the value of $\Gamma^{w}_{Na}$ for \ce{Na^+} changes from negative to positive
in a neutral POPC bilayer versus in a negatively charged bilayer with the composition 5\,PC:1\,PS.
%This means that at the given concentration the bilayer interface has a small preference to \ce{Na^+} cations compared to water molecules. 
Interestingly, this value is slightly decreased in the presence of an additional \ce{NaCl} concentration adding also \ce{Cl^-} anions, 
which are not present in the system when only counterions are used
(bottom resp. top plot in Fig.~\ref{fig:nacl-dens_PCPS}). 
The relative surface excess of calcium, $\Gamma^{w}_{Ca}$, also reflects
the higher affinity of the negatively charged PC:PS (5:1) membrane compared to a neutral POPC bilayer; 
its value for the latter is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$, 
while for the PC:PS mixture it is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable concentrations $C _{Ca} = 260\pm 30 \mathrm{mM} $.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_cacl.pdf} 
  \includegraphics[width=\figwidthsmall]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_cacl.pdf} 
  \caption{\label{fig:delta_ordPar_CaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer as a function of \ce{CaCl2} concentration 
    in bulk ($C_{ion}$) from simulations with different force fields and experiments at 298~K. \citep{roux90}
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  } 
  \todo{SAMULI: The forking of PS alpha carbon should not be put to zero with zero salt concentrations
    in this or in other figures. This is confusing expecially for the results with CaCl$_2$ where the forking
    seems to increase with CaCl$_2$, which is not correct. In NMRlipids IV, I have now put the lower order parameter
    to zero and kept the forking correct with all concentrations of ions.} \\
  \todo{SAMULI: Ion concentrations should be updated to the concentrations in water before solvating the lipids.} \\
  \todo{SAMULI: Error bars should be recalculated with the fixed code} \\
  \todo{SAMULI: Maybe we should have at least one more points with smaller concentration
    (maybe close to 100mM in solution before solvating lipids).} \\
\end{figure}

@@

\todo{SAMULI: The qualitative iprovement of PS headgroup order parameter response to bound Ca2+
  should be discussed. We should, however, point out that the response PS alpha carbon to bound
  Ca2+ is not excatly reproduced even by the ECC-POPS model.
}

\begin{figure}[htbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_ca_na_cl_wat_phos_models-compar_1-3_CaCl2-series.pdf}
  \caption{\label{fig:cacl-dens_PCPS} 
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along the normal of the membrane starting at the centre
    for the negatively charged membrane with the composition 5\,PC:1\,PS
    at various bulk concentrations of \ce{CaCl2} from simulations with ECC-lipids. 
    All profiles contain \ce{Na^+} counterions and an additional concentration of \ce{CaCl2}. 
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}},  
    the density profiles of~\ce{Cl^-} ions are divided by 2, and 
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively.  
  }
  \todo{SAMULI: Ion concentrations should be updated to the concentrations in water before solvating the lipids.} \\
  \todo{SAMULI: We should compare to the density profiles from Lipid17 simulations. These are also needed for NMRlipids IV.}
\end{figure}

@@

\todo{SAMULI: This paragraph in unclear. Should be clarified and the changes
  of headgroup orientations in PC and PS clearly discussed.
}
%%% discuss the table with Gammas and follow up with molecular details, where do cations bind? Support with the spatial density figure
The density profiles of ions suggest that
the dominant contribution to the binding of \ce{Ca^{2+}} to phospholipid bilayers
comes from the interactions with the phosphate groups in both POPC and POPS. 
%The population analysis also suggests 
%that the calcium cations prefer to reside in the phosphate region of the membrane. 
%Such a finding is further validated with a more detailed analysis using Markov state modeling (MSM) \citep{Pande_MSM_paper} \todoi{Add papers reviewing MSM, e.g. recent Pande's paper.}. 
%The set of states of a calcium cation 
%encompassed all possible combinations of up to three surrounding lipids. 
%In the case of PC, we used only the phosphate moitety, which forms the dominant contribution for calcium binding.
%For PS we also distinguished configurations in which calcium interacts with the carboxylate moiety. 
%All possible combinations of such states were used to build a Markov model at a lag time $25\,\mathrm{ns}$ using pyEMMA code by \citet{pyemma} \todoi{add citation for pyemma}. 
%The resulting MSM was validated using Chapman-Kolmogorov test \citep{FrankNoe_papers_MSM} \todoi{Add citation for Noe's papers on MSM, especially Chapman Kolmogorov test.}. 
%The stationary distribution of the states reveals a strong preference of the calcium cations to reside in the phosphate region of the mixed PC-PS membrane. 
%When interacting with PS, configurations with the phosphate moiety from either PC or PS dominate the population.
%Moreover, states containing interactions with the carboxylate group in PS 
%bear higher probability when interacting also with the phosphate group of the same lipid or other lipids. 
This is also reflected in the shift of the mean orientation 
of the \ce{COO^-} group from $62^\circ$ to $73^\circ$ (420~mM \ce{CaCl2})
which was measured as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid. 
The interactions of the carboxylate group in PS with calcium and other phosphate groups
shed light into the qualitatively different response of the head group order parameters $\alpha$ and $\beta$ in PS compared to PC 
(see Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_CaCl_PCPS}). 
We find that the complex response of the head group order parameters of POPS 
is affected by the conformational changes of the carboxylate group,
which is attracted more towards the phosphate region, 
where the calcium cations dominantly bind. 
In line with the experimental work by \citet{browning80},
this is also very likely the reason, 
why the magnitude of the P-N vector change in POPS is diminished compared to POPC, 
which is not restrained by an additional cation binding group like \ce{COO^-} in POPS.
